**Microbiota predict _Clostridium difficile_ severity in germ-free mice colonized with human feces**

### 2016-11-13 submission/Manuscript.Rmd
almost done with results

---
title: "**Microbiota predict _Clostridium difficile_ severity in germ-free mice colonized with human feces**"
[//]: # (bibliography: references.bib)
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

### Introduction
Clostridium difficile is an opportunistic pathogen of the human lower gastrointestinal tract. Disruption of the native microbial community of the gut by antibiotics is the most common risk factor for development of C. difficile infection (CDI) (1). C. difficile is a spore-forming bacteria and can persist on abiotic surfaces and is not readily killed by ethanol-based hand-sanitizers, putting hospital patients particularly at risk. Indeed, ~12% of hospital acquired infections in the United States are due to C. difficile and result in up to 15,000 deaths annually (2).

**need to add a paragraph about colonization resistence**

Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility (3, 4). This model provides a convenient way to study C. difficile pathogenesis and virulence factors. Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of CDI. While informative, these studies are somewhat removed from human disease because they only examine mouse-associated microbial communities.

Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between C. difficile and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of C. difficile and examination of host immune response to CDI (10). Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11), (12). Pigs have a longer digestive tract with components more similar to humans than mice and are typically infected by strains typical in human infection (13). However, the murine and porcine microbiomes typically do not resemble those of the human gut.

The power of the gnotobiotic models to study CDI has been further realized by first inoculating germ-free mice and piglets with human stool microbes. In one study, germ-free piglets were acutely colonized with human feces for one week and then treated with tigecycline. After challenge with C. difficile none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11). Further, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14). These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by C. difficile (14).  While informative, these studies were limited in their use of only one human donor as input inoculum. In order to best understand the impact of C. difficile pathogenesis on human disease, we must have a laboratory model that allows for study of a variety of human-derived microbiomes.

To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors. We then characterized human-associated microbiome response to C. difficile challenge. Additionally, the use of machine-learning models allowed us to build a predictive model that classified “at-risk” microbiomes prior to infection with C. difficile. These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotics and that study of mice colonized with human feces provides a range of clinical outcomes.

@@

@@

**Microbes present in the gut prior to infection are predictive of C. difficile colonization levels**

**Results to be written**
1.	Microbes present in the gut prior to infection are predictive of C. difficile CFU and severity 
a.	Figure 3: Random forest to predict CFU, predictive OTUs 
b.	Figure 4: Random Forest predicts CDI severity, predictive OTUs
2.	Propensity for severe CDI is community-dependent and strain-independent
a.	Figure 5:  Infection of mice with different C. difficile strains.

>>>>

---
title: "**Microbiota predict _Clostridium difficile_ severity in germ-free mice colonized with human feces**"
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

### Introduction
_Clostridium difficile_ is an opportunistic pathogen of the human lower gastrointestinal tract. _C. difficile_ forms spores that can persist on abiotic surfaces and are not readily killed by ethanol-based hand-sanitizers, putting hospital patients particularly at risk. Indeed, ~12% of hospital acquired infections in the United States are due to _C. difficile_ and result in up to 15,000 deaths annually (2). Disruption of the native microbial community is the most common risk factor for development of _C. difficile_ infection (CDI) (1). Antibiotic use and inflammatory bowel diseases are associated with loss of colonization resistance to _C. difficile_ through the loss of potentially protective bacterial families such as _Barnesiella_ and _Lachnospiraceae_ in both mouse models and human association experiments (cite alyx 2014 and vincent 2013). The composition of the community is clearly important for the acquisition, resistance and treatment of _C. difficile_ infection, as giving patients a healthy fecal microbiome transplant is the most effective treatment for this disease (cite Anna 2014). The precise mechanisms of colonization resistance and restoration of a healthy community are yet to be discovered.

Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility (3, 4). This model provides a convenient way to study _C. difficile_ pathogenesis and virulence factors. Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of _C. difficile_ infection. While informative, these studies are somewhat removed from human disease because they only examine mouse-associated microbial communities.

Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between _C. difficile_ and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of _C. difficile_ and examination of host immune response to CDI (10). Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11), (12). Pigs have a longer digestive tract with components more similar to humans than mice and are typically infected by strains typical in human infection (13). However, the murine and porcine microbiomes typically do not resemble those of the human gut.

The power of the gnotobiotic models to study CDI has been further realized by first inoculating germ-free mice and piglets with human stool microbes. In one study, germ-free piglets were acutely colonized with human feces for one week and then treated with tigecycline. After challenge with C. difficile none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11). In another study, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14). These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by _C. difficile_ (14).  While informative, these studies were limited in their use of only one human donor as input inoculum. In order to best understand the impact of _C. difficile_ pathogenesis on human disease, we must have a laboratory model that allows for study of a variety of human-derived microbiomes.

Here we designed a laboratory model that allows for studying and modeling microbial coumminity interactions with _C. difficile_ infection in mice with human derived microbiomes. To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors. We then characterized human-associated microbiome response to C. difficile challenge. Additionally, the use of machine-learning models allowed us to build a predictive model that classified “at-risk” microbiomes prior to infection with C. difficile. These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotics and that study of mice colonized with human feces provides a range of clinical outcomes.

@@

**need to put in a line to reference supplemental table with RA in it**

@@

**Microbes present in the gut prior to infection are predictive of _C. difficile_ colonization levels** Previous work in our group has demonstrated that the microbes present in the mouse microbiome prior to _C. difficile_ challenge can predict future colonization levels (cite Alyx). To determine if the human-associated microbiome in the mouse could similarly predict _C. difficile_ colonization we employed a Random Forest machine learning algorithm similar to the one used in previous studies. Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model. We found that the model explained X% of the variance in colonization levels of _C. difficile_ (Fig 3A). By refining the model to include only the top X predictive OTUs, the model explained X% of the variance. Partial dependency plots of the top 6 OTUs predictive of _C. difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B). Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_ (Fig 2A, 3B). These results confirm that as in mouse microbiome studies, the human-associated gut community prior to challenge is predictive of _C. difficile_ colonization levels.

**The microbiome predicts _C. difficile_ severity.** As _C. difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0. Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease. We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med). The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B). The populations that were most commonly found in mice that succumbed to _C. difficile_ infection were members of the _Clostridiales_, _Bacilliales_, _Streptococcus_ (Lactobacilles?) and _Bacteriodes_. (need more OTUs here, or number that were used for model and then group by rank). The AUC was calculated using the Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using 10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not significantly different from the OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes.

**Results to be written**

b.	Figure 4: Random Forest predicts CDI severity, predictive OTUs
2.	Propensity for severe CDI is community-dependent and strain-independent
a.	Figure 5:  Infection of mice with different C. difficile strains.
